Boosting Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models by Hamiltonian Scale Separation

### 2015-10-23 TeXFiles/Manuscript_Albert_Ulzega_Stoop.tex
improvement of tex file, new citations added

\title{Boosting Bayesian Parameter Inference of Nonlinear Stochastic Differential Equation Models by  Hamilton Equation Scale Separation }

@@

\begin{abstract}
For the modeling of dynamical systems characterized by time-series, we propose a novel very efficient approach for generating posterior parameter distributions. The algorithm is inspired by re-interpreting the posterior distribution as a partition function of a
%1D
statistical mechanics system. To arrive at distribution samples, we employ a Hamiltonian Monte Carlo approach combined with multiple time-scale integration.
At least for 1D problems, generic re-parametrization allows us to decouple the harmonic modes between measurement points and solve their dynamics analytically.
%Generic re-parametrization helps us to decouple the harmonic modes between measurement points, which leads to an analytical solution of the node dynamics.
Our approach is applicable to a wide range of inference problems and is highly parallelizable.

@@

Modeling a dynamical process starts with a basic model of the process, which usually is obtained from more or less deep insight into the nature of the process. The next step then is the determination of the parameters of the model, based on observed data, which is generally a highly nontrivial task, in particular when complex behavior of such systems needs to be predicted, or when the measurements are noisy.
A minimal example is a perceptron [XXX], the basic element of a neuronal network, that predicts the double cosine value associated with the input of the corresponding sine function plus the sine's value at a fixed earlier time. While this task can easily be achieved for clean data using gradient descent learning, for noisy input data, this is largely impossible, as noise cannot be learned. The result is a distribution of potential parameter values (Fig.~(\ref{neural})).
\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\textwidth]{neuron.pdf}
    \caption{General problem setting: Parameter estimation (synaptic weights $w_1$,$w_2$) of a perceptron, from a) noiseless, b) noisy data (noise sampled from a flat distribution over the  interval $[-0.2,0.2]$). Whereas for noiseless data the estimates perfectly converge, for noisy data the estimates  the system attempts to also include the noise, leading to a nontrivial distribution of the parameter estimates, and rendering the extraction of the optimal parameters a nontrivial task. Red circles: location of the optimal parameters in the noiseless case.}
    \label{neural}
\end{figure}
If the physics underlying the evolution of the system follows a differential equation,
the search for the optimal parameters and their typical distribution is called Bayesian inference of a stochastic differential equation (SDE).

In Bayesian statistics, knowledge about parameters is expressed by probability distributions and learning is implemented as an update rule on these distributions [REF].
If a constant noise term is added to the output of a (reasonably fast) deterministic model, such as in the perceptron example above, Bayesian inference is straightforward.
If noise enters the formulation of the model equations, however, Bayesian inference quickly becomes computationally very expensive due to the high dimensional integrals that have to be evaluated.

@@

%In our approach we remedy these difficulties.
%We start by re-interpreting the posterior distribution as the partition function of a statistical mechanical system, of which we simulate the time evolution.
These difficulties can be remedied with a reinterpretation of the Bayesian posterior distribution as the partition function of a statistical mechanical system and by simulating the dynamics of the latter.
After discretizing the time of the original problem, we are led to the statistical mechanics of a polymer %-like problem
with harmonic bonds in an exterior potential. In this framework, the measurements are interpreted as an additional exterior potential that acts only on the polymer's `measurement beads', and model parameters are interpreted as additional degrees of freedom coupling to all the beads of the polymer.
To simulate the dynamics of this system, we apply the Hamiltonian Monte Carlo (HMC) algorithm \cite{duane_1987}, which combines Molecular Dynamics (MD) \cite{alder_1959_MD, rahman_1964_MD} with the Metropolis algorithm  \cite{metropolis_1953}.
Compared to traditional methods, the use of the Hamiltonian approach achieves much higher acceptance rates since data points are already used for the suggestion of new parameters, and thus model realizations incompatible with the data are never considered. The drawback is that the model equations need to be known and derivatives have to be calculated.

HMC requires two sets of parameters to be tuned: (i) the parameters that define the kinetic energy of the statistical mechanical system and (ii) those parameters that define the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm. Efficiency of HMC algorithms can be gained if the kinetic term is made dependent on the configuration geometry of the statistical mechanical system. If the Riemann geometry of the parameter space of statistical models is taken into account, then the simulated search of paths across this manifold samples the target density in an utmost efficient way
    \cite{girolami_2011_HMC}. Unfortunately, this procedure is both demanding and computationally costly,  depending strongly on the quality of the space's extracted geometry.

@@

Here we
%therefore
explore a computationally simpler approach, which, to our knowledge, has never been applied in the context of Bayesian inference before.
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happen on very different time scales. This suggests a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
We will show that for 1D SDEs we can always find a parametrization, which decouples the harmonic modes in between measurement points from both the measurement points and the model parameters and allows for an analytical time-saving solution of the fastest part of the dynamics.
Whilst for higher dimensional problems it is not always possible to solve part of the dynamics analytically, we believe that scale separation alone will render many SDEs amenable to a full-fledged Bayesian inference with time-series.
In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretizaion points.
%Our approach can be combined with the approach in \cite{girolami_2011_HMC}.

@@

\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input, $\gamma$ is the noise strength, and $\eta(t)$ indicates the white noise property, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
For constant input $r(t) = r_0$ and in the long-time limit, the mean of $S(t)$ converges to the equilibrium solution of the unperturbed $(\gamma = 0)$ system, $S_{eq} = Kr_{0}$ (cf. end of section).

@@

While this setting was motivated by a popular hydrological model \cite{breinholt_2011_SDE}, it is by no means restricted to this context. By means of the transformation $S(t) = 1/n(t)$, Eq.~(\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and has extensively been studied from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.
In this setting, the input $r(t)$ is a smooth and nowhere vanishing function. We assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors,

\begin{equation}\label{data}
  \ln \left( y_s \right)
  =
  \ln \left( \frac{S(t_s)}{K} \right)
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
\end{equation}
where the $\epsilon_i$ are uncorrelated standard normal errors.
For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known, so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}) in the Bayesian sense, where knowledge about parameters, $\bt$, is expressed in terms of probability distributions.
We start our treatise by assuming that we have prior knowledge in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, believed to be a realization of the model.
The posterior knowledge, combining prior knowledge with the one acquired from data, is calculated by means of
\begin{equation}\label{Bayes}
  f_{\mbox{\small post}}(\bt|\vc y)
  =
  \frac{f_{\mbox{\small prior}}(\bt)L(\vc y|\bt)}
  {\int f_{\mbox{\small prior}}(\bt')L(\vc y|\bt')d\bt'}\,,
\end{equation}
where $L(\vc y|\bt)$ is the probability distribution for model outputs given model parameters, evaluated at the measured data (the so-called {\em likelihood function}).
We posit that parameters must be non-negative; otherwise we assume no additional prior knowledge (i.e., the posterior is ‘data-driven’ [XX]).

@@

Because this will permit to do part of the integration to be performed later analytically, it is important to arrive at model equations where the noise term neither depends on the state variables, nor on the parameters to be inferred. In a one dimensional model this can always be achieved through re-parametrization [XXX]. The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ if it was in an initial state $q_0$ at time $t = 0$, is expressed as a {\em path-integral} as

@@

Instead of undertaking an often prohibitive numerical computation of the path integral, we now apply HMC to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
For efficiently drawing parameter samples from (\ref{posterior_pathint}), we interpret the latter as the partition function of a 1D statistical mechanical system and simulate its dynamics employing the HMC algorithm \cite{duane_1987}. (This is always possible and does not depend on the nature of the SDE we consider.) The model parameters  $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$. Each degree of freedom, $q(t)$ and $\bt$, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$ respectively, so that the system is defined by the Hamiltonian

@@

Physically, the discretized Hamiltonian can be identified with a classical polymer chain of $N$ beads with harmonic bonds between neighboring beads in an external field. The latter consists of two parts, a field that results from the dynamics of the original Eq.~(\ref{sde}) and is felt by all the beads, and a field that results from the measurements and is felt by the measurement beads only. The masses $m_i$ and $m_{\alpha}$ are tunable parameters of the algorithm. Since measurement beads are constrained more than intermediate beads, we will assign larger masses to the former. Fig. (\ref{fig:polymer}) shows a typical realization of the dynamics of the polymer. The physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced as the vertical axis. Measurement beads only move within the measurement uncertainty, whilst the intermediate beads explore much larger regions of phase space.

@@

We have thus reduced the original Bayesian inference problem to simulating the dynamics of a linear polymer (cf. Fig.~(\ref{fig:polymer})). Each state of this fictitious molecule corresponds to a well-defined configuration in the original phase space, characterized by a set of system variables $\left\{q_i\right\}_{i=1,\dots,N}$ and a parameter vector $\bt$. It is now essential to note that potential (\ref{Vdisc}) contains terms of distinct scaling in the potentially large numbers $N$ and $n$, that refer to dynamics on distinct time-scales. In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part, and for resolving its dynamics, brute force numerical integration of Hamilton's equations in step ($ii$) of the HMC algorithm would require a very small discretization time-step.

@@

In the following we restrict ourselves to regime $(ii)$, where the number of measurements $n$ is assumed to be not too large and/or the measurement error $\sigma$ to be not too small (the generalization of the method to the other regimes would, however, be straightforward). In this regime we may simply separate the harmonic part of the action for the staging beads from the rest and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
\end{equation}
For obtaining reversible integrators, we define the Liouville operators
$ iL_N=\{\cdot\,,\,\mathcal H_N\}$, $iL'=\{\cdot\,,\,\mathcal H'\}$,
where $\{\cdot\,,\,\cdot\}$ denote the Poisson brackets that apply to functions on the phase space.
Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
\begin{equation}\label{propagator}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(\tau^3/P^{2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
Here, the outer propagator $\exp[iL_N(\Delta \tau/2)]$ reflects much faster dynamics than the inner one.
However, thanks to our re-parametrization, it describes the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m'/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The outer propagator becomes
\begin{align}
  u_{(s-1)j+k}(\Delta\tau/2)
  &=
  u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,
  \\
  p_{(s-1)j+k}(\Delta\tau/2)
  &=
  p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{align}
for $s=1,\dots,n$ and $k=2,\dots,j$.
This elegant `analytical' solution is one main boosting part of our algorithm.
For the inner propagator, we employ the time-reversible and symplecticity preserving velocity Verlet algorithm \cite{swope_1982_verlet}, which leads for the boundary beads to
\begin{align}
  u_{(s-1)j+1}(\Delta\tau)
  &= u_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{M} p_{(s-1)j+1}(0)
  +
  \frac{\Delta \tau^2}{2M}
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
  p_{(s-1)j+1}(\Delta\tau)
  &= p_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )\,,
\end{align}
with $s=1,\dots,n+1$ and where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
Analogous equations emerge for the model parameters $\bt$ and their momenta ${\pmb\pi}$, by exchanging $u_{…}$ by $\theta_{…}$ and $p_{…}$ by $\pi_{…}$, along with the corresponding masses, respectively.
For the staging beads
%, as yet another benefit of the approach,
only the momenta need to be updated (because the associated kinetic term is not part of $\mathcal H'$, but of $\mathcal H_N$).
Thus, with $s=1,\dots,n$ and $k=2,\dots,j$ as before,
\begin{align}
  p_{(s-1)j+k}(\Delta\tau)
  &=
  p_{(s-1)j+k}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )
  \,.
\end{align}

@@

We presented a novel, extremely efficient and versatile approach for SDE parameter estimation, based on observed data.
Our algorithm derives its strenght from a translation of the problem of generating posterior parameter samples to the problem of simulating the dynamics of a statistical mechanical system.
The main efficiency gain of our algorithm, compared to the one suggested in Ref. \cite{girolami_2011_HMC}, stems from an exploitation of the fact that this dynamics generically happens on very different time scales.
Furthermore, at least for 1D systems, our approach also allows for an analytical, and therefore computationally efficient, integration of the fastest part of the dynamics.
The algorithm suggested in Ref. \cite{girolami_2011_HMC} locally adapts the mass matrix to the curvature of the potential, whereas we use a constant diagonal mass matrix, with reasonable choices of masses, heavier for the measurement beads, lighter for the discretization (staging) beads.
Whilst the former approach leads to some computational overhead, as second derivatives have to be calculated and implicit equations have to be solved, it is likely to be beneficial in cases with strongly varying curvature.
Our approach, however, is to a large extent complementary to and could be combined with the one in \cite{girolami_2011_HMC}.

@@

The structure of our algorithm is well suited to parallelization, which is important in particular if we deal with a high number of measurements (e.g., some tens of thousands data points). The algorithm can easily be adapted to other inference problems, such as higher dimensional SDE, and SDE coupled to ODE. These adaptations and extensions will, however, be addressed in future works.

>>>>

\title{Boosting Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models by Hamiltonian Scale Separation}

@@

\begin{abstract}
We propose a novel very efficient approach for generating posterior parameter distributions, for the modeling of dynamical systems characterized by time-series. The algorithm is inspired by re-interpreting the posterior distribution as a partition function of a
%1D
statistical mechanics system. To arrive at distribution samples, we employ a Hamiltonian Monte Carlo approach combined with multiple time-scale integration.
At least for 1D problems, generic re-parametrization allows us to decouple the harmonic modes between measurement points and solve their dynamics analytically.
%Generic re-parametrization helps us to decouple the harmonic modes between measurement points, which leads to an analytical solution of the node dynamics.
Our approach is applicable to a wide range of inference problems and is highly parallelizable.

@@

Modeling a dynamical process starts with a basic model that is usually obtained from more or less deep insight into the nature of the process. The next step then is the determination of the parameters of the model, based on observed data, which is generally a highly nontrivial task, in particular when complex behavior of such systems needs to be predicted, or when the measurements are noisy.
A minimal example is a perceptron [XXX], the basic element of a neuronal network, that predicts the double cosine value associated with the input of the corresponding sine function plus the sine's value at a fixed earlier time. While this task can easily be achieved for clean data using gradient descent learning, for noisy input data, this is largely impossible, as noise cannot be learned. The result is a distribution of potential parameter values (Fig.~(\ref{neural})).
\begin{figure}[h]
    \centering
    \includegraphics[width=1.0\textwidth]{neuron.pdf}
    \caption{General problem setting: Parameter estimation (synaptic weights $w_1$,$w_2$) of a perceptron, from a) noiseless, b) noisy data (noise sampled from a flat distribution over the  interval $[-0.2,0.2]$). Whereas for noiseless data the estimates perfectly converge, for noisy data the estimates  the system attempts to also include the noise, leading to a nontrivial distribution of the parameter estimates, and rendering the extraction of the optimal parameters a nontrivial task. Open circles: location of the optimal parameters in the noiseless case.}
    \label{neural}
\end{figure}

In Bayesian statistics, knowledge about parameters is expressed by probability distributions and learning is implemented as an update rule on these distributions [REF].
If a constant noise term is added to the output of a deterministic model, such as in the perceptron example above, Bayesian inference is straightforward.
If noise enters the formulation of the model equations, however, Bayesian inference all of a sudden becomes computationally very expensive.

@@

%In our approach we remedy these difficulties.
%We start by re-interpreting the posterior distribution as the partition function of a statistical mechanical system, of which we simulate the time evolution.
These difficulties can be remedied with a reinterpretation of the Bayesian posterior distribution as the partition function of a statistical mechanics system and by simulating the dynamics of the latter.
After discretizing the time of the original problem, we are led to the statistical mechanics of a polymer %-like problem
with harmonic bonds in an exterior potential. In this framework, the measurements are interpreted as an additional exterior potential that acts only on the polymer's `measurement beads', and model parameters are interpreted as additional degrees of freedom coupling to all the beads of the polymer.
To simulate the dynamics of this system, we apply the Hamiltonian Monte Carlo (HMC) algorithm \cite{duane_1987}, which combines Molecular Dynamics (MD) \cite{alder_1959_MD, rahman_1964_MD} with the Metropolis algorithm  \cite{metropolis_1953}.
Compared to traditional methods, the use of the Hamiltonian approach achieves much higher acceptance rates since data points are already used for the suggestion of new parameters, and thus model realizations incompatible with the data are never considered. The drawback is that the model equations need to be known and derivatives have to be calculated.

HMC requires two sets of parameters to be tuned: (i) the parameters that define the kinetic energy of the statistical mechanics system and (ii) the parameters that define the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm. Efficiency of HMC algorithms can be gained if the kinetic term is made dependent on the configuration geometry of the statistical mechanics system. If the Riemann geometry of the parameter space of statistical models is taken into account, then the simulated search of paths across this manifold samples the target density in an utmost efficient way
    \cite{girolami_2011_HMC}. Unfortunately, this procedure is both demanding and computationally costly,  depending strongly on the quality of the space's extracted geometry.

@@

Here we
%therefore
explore a computationally simpler approach, which, to our knowledge, has never been applied in the context of Bayesian inference before.
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanics system happen on very different time scales. This suggests a multiple time scale integration technique for the simulation of the statistical mechanics system \cite{tuckerman_1993}.
We will show that for 1D SDEs we can always find a parametrization, which decouples the harmonic modes in between measurement points from both the measurement points and the model parameters and allows for an analytical time-saving solution of the fastest part of the dynamics.
Whilst for higher dimensional problems it is not always possible to solve part of the dynamics analytically, we believe that scale separation alone will render many SDEs amenable to a full-fledged Bayesian inference with time-series.
In fact, scale separation appears to be a generic feature if the dynamics of the SDE requires a large number of discretizaion points.
%Our approach can be combined with the approach in \cite{girolami_2011_HMC}.

@@

\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input, $K$ denotes the retention time,$\gamma$ is the noise strength, and $\eta(t)$ indicates the white noise property, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
For constant input $r(t) = r_0$ and in the long-time limit, the mean of $S(t)$ converges to the equilibrium solution of the unperturbed $(\gamma = 0)$ system, $S_{eq} = Kr_{0}$ (cf. end of section).

@@

While this setting was motivated by a popular hydrological model \cite{breinholt_2011_SDE, livina_2003_dischargeFluctuations}, it is by no means restricted to this context. By means of the transformation $S(t) = 1/n(t)$, Eq.~(\ref{sde}) turns into a model that has been suggested, e.g., as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and has extensively been studied from a theoretical point of view \cite{schenzle_1979_multStochProc, fujisaka_1986_intermittency}.
In this setting, the input $r(t)$ is a smooth and nowhere vanishing function. We assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors,

\begin{equation}\label{data}
  \ln \left( y_s \right)
  =
  \ln \left( \frac{S(t_s)}{K} \right)
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
\end{equation}
where the $\epsilon_i$ are uncorrelated standard normal errors.
For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known, so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}) in the Bayesian sense, where knowledge about parameters is expressed in terms of probability distributions.
We start our treatise by assuming that we have prior knowledge about a parameter vector, $\bt$, in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, believed to be a realization of the model.
The posterior knowledge, combining prior knowledge with the one acquired from data, is calculated by means of
\begin{equation}\label{Bayes}
  f_{\mbox{\small post}}(\bt|\vc y)
  =
  \frac{f_{\mbox{\small prior}}(\bt)L(\vc y|\bt)}
  {\int f_{\mbox{\small prior}}(\bt')L(\vc y|\bt')d\bt'}\,,
\end{equation}
where $L(\vc y|\bt)$ is the probability distribution for model outputs given model parameters, evaluated at the measured data (the so-called {\em likelihood function}).
We posit that parameters must be non-negative; otherwise we assume no additional prior knowledge (i.e., the posterior is ‘data-driven’ [XX]).

@@

At this point, it is important to arrive at model equations where the noise term neither depends on the state variables, nor on the parameters to be inferred.
This will permit us to perform an integration that is needed in a later step analytically.
In a one dimensional model this can always be achieved through re-parametrization (see, e.g., chapter 5 in Ref. \cite{risken_1989_FockerPlanck}). The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ if it was in an initial state $q_0$ at time $t = 0$, is expressed as a {\em path-integral} as

@@

Instead of undertaking an often prohibitive numerical computation of the path integral, we now apply HMC to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.
For efficiently drawing parameter samples from (\ref{posterior_pathint}), we interpret the latter as the partition function of a 1D statistical mechanics system and simulate its dynamics employing the HMC algorithm \cite{duane_1987}. The model parameters  $\bt$ are interpreted as additional dynamical degrees of freedom coupling to the system variables $q(t)$. Each degree of freedom, $q(t)$ and $\bt$, is paired with a conjugate variable, $p(t)$ and ${\pmb\pi}$ respectively, so that the system is defined by the Hamiltonian

@@

Physically, the discretized Hamiltonian can be identified with a classical polymer chain of $N$ beads with harmonic bonds between neighboring beads in an external field.
The latter consists of two parts, a field that results from the measurements and is felt by the measurement beads only (last term on the r.h.s. of eq. \ref{Vdisc}), and a field that results from the dynamics of the original Eq.~(\ref{sde}) and is felt by all the beads.
The masses $m_i$ and $m_{\alpha}$ are tunable parameters of the algorithm. Since measurement beads are constrained more than intermediate beads, we will assign larger masses to the former. Fig. (\ref{fig:polymer}) shows a typical realization of the dynamics of the polymer. The physical time (horizontal axis) is interpreted as a spatial dimension while a fictitious simulation time is introduced as the vertical axis. Measurement beads only move within the measurement uncertainty, whilst the intermediate beads explore much larger regions of phase space.

@@

We have thus reduced the original Bayesian inference problem to simulating the dynamics of a linear polymer (cf. Fig.~(\ref{fig:polymer})). Each state of this fictitious molecule corresponds to a well-defined configuration in the original phase space, characterized by a set of system variables $\left\{q_i\right\}_{i=1,\dots,N}$ and a parameter vector $\bt$. It is now essential to note that potential (\ref{Vdisc}) contains terms of distinct scaling in the potentially large numbers $N$ and $n$, that refer to dynamics on distinct time-scales. In particular, for large $N$, $V(q,\bt)$ is dominated by its harmonic part, and to resolve its dynamics, brute force numerical integration of Hamilton's equations in step ($ii$) of the HMC algorithm would require a very small discretization time-step.

@@

In the following we restrict ourselves to regime $(ii)$, where the number of measurements $n$ is assumed to be not too large and/or the measurement error $\sigma$ to be not too small (the generalization of the method to the other regimes would, however, be straightforward). In this regime we may simply separate the harmonic part of the action for the staging beads from the rest and write
\begin{equation}
  \mathcal{H}_{\mbox{\tiny HMC}}=\mathcal H_N + \mathcal H'\,.
\end{equation}
For obtaining reversible integrators, we define the Liouville operators
$ iL_N=\{\cdot\,,\,\mathcal H_N\}$, $iL'=\{\cdot\,,\,\mathcal H'\}$,
where $\{\cdot\,,\,\cdot\}$ denote the Poisson brackets that apply to functions on the phase space.
Trotter's formula \cite{trotter_1959} allows us to write the Hamiltonian propagator as
\begin{equation}\label{propagator}
  e^{i(L_N+L')\tau}
  =
  (e^{iL_N(\Delta\tau/2)}e^{iL'\Delta\tau}e^{iL_N(\Delta\tau/2)})^P
  +
  \mathcal O(\tau^3/P^{2})\,,
\end{equation}
for $\tau =P\Delta \tau$.
Here, the outer propagator $\exp[iL_N(\Delta \tau/2)]$ reflects much faster dynamics than the inner one.
However, thanks to our re-parametrization, it describes the dynamics of uncoupled harmonic oscillators, which we can readily solve.
Masses and frequencies of the oscillators are derived from (\ref{H_N}) as
\begin{equation}
  m=m'/\Delta t\,,\quad
  \omega_k=\sqrt{\frac{Nk}{(k-1)m}}\,.
\end{equation}
The outer propagator becomes
\begin{align}
  u_{(s-1)j+k}(\Delta\tau/2)
  &=
  u_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  +
  \frac{p_{(s-1)j+k}(0)}{m\omega_k}\sin(\omega_k\Delta\tau/2)\,,
  \\
  p_{(s-1)j+k}(\Delta\tau/2)
  &=
  p_{(s-1)j+k}(0)\cos(\omega_k\Delta\tau/2)
  -
  m\omega_k u_{(s-1)j+k}(0) \sin(\omega_k\Delta\tau/2)\,,
\end{align}
for $s=1,\dots,n$ and $k=2,\dots,j$.
This elegant `analytical' solution is one main boosting part of our algorithm.
For the inner propagator, we employ the time-reversible and volume preserving velocity Verlet algorithm \cite{swope_1982_verlet}, which leads for the boundary beads to
\begin{align}
  u_{(s-1)j+1}(\Delta\tau)
  &= u_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{M} p_{(s-1)j+1}(0)
  +
  \frac{\Delta \tau^2}{2M}
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]\,,\\
  p_{(s-1)j+1}(\Delta\tau)
  &= p_{(s-1)j+1}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+1}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+1}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )\,,
\end{align}
with $s=1,\dots,n+1$ and where $F_i[\vc u,{\pmb\theta}]$ denotes the partial derivative of $\mathcal H'[\vc u,{\pmb\theta}]$ w.r.t. $u_i$.
Analogous equations emerge for the model parameters $\bt$ and their momenta ${\pmb\pi}$, by exchanging $u_{…}$ by $\theta_{…}$ and $p_{…}$ by $\pi_{…}$, along with the corresponding masses, respectively.
For the staging beads
%, as yet another benefit of the approach,
only the momenta need to be updated (because the associated kinetic term is not part of $\mathcal H'$, but of $\mathcal H_N$).
Thus, with $s=1,\dots,n$ and $k=2,\dots,j$ as before,
\begin{align}
  p_{(s-1)j+k}(\Delta\tau)
  &=
  p_{(s-1)j+k}(0)
  +
  \frac{\Delta\tau}{2}
  (
  F_{(s-1)j+k}[\vc u(0),{\pmb\theta}(0)]
  +
  F_{(s-1)j+k}[\vc u(\Delta\tau),{\pmb\theta}(\Delta\tau)]
  )
  \,.
\end{align}

@@

We presented a novel, extremely efficient and versatile approach for data-based SDE parameter estimation.
Our algorithm obtains its strength from translating the problem of generating posterior parameter samples into the problem of simulating the dynamics of a statistical mechanics system.
The main novelty in our algorithm is the exploitation of the fact that this dynamics generically happens on very different time scales.
Furthermore, at least for 1D systems, our approach also allows for an analytical, and therefore computationally efficient, integration of the fastest part of the dynamics.

Our algorithms works with a constant diagonal mass matrix, with reasonable choices of masses - heavier for the measurement beads, lighter for the discretization (staging) beads.
We expect this to work well in most application cases.
However, if the curvature of the potential varies strongly, it might be beneficial to adapt the mass matrix to the local curvature as suggested in Ref. \cite{girolami_2011_HMC}.
This, in turn, comes with some computational overhead, as second derivatives have to be calculated and implicit equations have to be solved.
Furthermore, we will no longer be able to solve part of the dynamics analytically.
Nonetheless, for certain applications, a combination of the scale separation method proposed in this paper with a local adaptation of the mass matrix from Ref. \cite{girolami_2011_HMC} might  be the most efficient approach.

@@

The structure of our algorithm is well suited to parallelization, which is important in particular if we deal with a high number of measurements. The algorithm can easily be adapted to other inference problems, such as higher dimensional SDE, and SDE coupled to ODE. These adaptations and extensions will, however, be addressed in future works.
